\bf Drift Removal for Time Series Data Using Quantile Trend Filtering

### 2018-12-07 Manuscript/simulation_notes.tex
update simulation notes and figures

We compare the performance of our quantile trend filtering method with the three previously published methods using designs proposed by \cite{Racine2017}. The methods compared are

@@

	\item \texttt{detrendr\_SIC}: Our method where we minimize
	$\sum_i\rho_{\tau}(y_i - \theta_i) + \lambda||D\theta||_1$ and $\lambda$ is chosen using SIC from above. A single value of $\lambda$ was chosen by scaling and summing SIC values across all quantiles. 
	\item \texttt{detrendr\_valid}: Our method where lambda is chosen by leaving out every 5th observation as a validation data set and evaluating the check loss function on the validation data.
	\item \texttt{detrendr\_eBIC}:   The traditional BIC is given by 
	\begin{equation}
	\mbox{BIC}(s) = -2\log(L\{\hat{\theta}(s)\}) + \nu(s)\log n 
	\end{equation}	
	where $\theta(s)$ is the parameter $\theta$ with those components outside $s$ being set to 0, and $\nu(s)$ is the number of components in $s$. If we assume an asymmetric Laplace likelihood $L(y|\theta) = \left(\frac{\tau^n(1-\tau)}{\sigma}\right)^n\exp\left\{-\sum_i\rho_\tau(\frac{y_i - \theta_i}{\sigma})\right\}$ and the number of non-zero elements of $D\theta$ as $df$
	\begin{equation}
	\mbox{BIC}(df) = 2\sum_i\frac{1}{\sigma}\rho_{\tau}(y_i-\theta_i) + df\log n
	\end{equation} 
	We can choose and $\sigma>0$ and have found empirically that $\sigma =  \frac{1-|1-2\tau|}{2}$ produces stable estimates. \cite{chen2008} proposed the extended BIC for large parameter spaces 
	\begin{equation}
	BIC_{\gamma}(s) = -2\log(L\{\hat{\theta}(s)\}) + \nu(s)\log n  + 2\gamma\log{P\choose j}
	\end{equation}
	where $P$ is the total number of possible parameters and $j$ is the number of parameters included in given model.

@@

Three simulation designs from \cite{Racine2017} were considered. For all designs $X_i$ was generated as a uniformly spaced sequence in $[0,1]$ and the response $Y$ was generated as 
$$Y_i = sin(2\pi x_i) + \epsilon_i(x_i)$$
The three error distributions considered were 
\begin{itemize}
 \item Gaussian: $\epsilon_i(x_i) \sim N\left(0, \left(\frac{1+x_i^2}{4}\right)^2\right)$
 \item Beta: $\epsilon_i \sim Beta(1, 11-10x_i)$
\item Mixed normal: $\epsilon_i$ is simulated from a mixture of $N(-1,1)$ and  $N(1,1)$ with mixing probability $x_i$.
\end{itemize}
\begin{figure}
\caption{Simulated data with true quantiles $\tau \in \{0.01, 0.05, 0.25, 0.5, .75, 0.95, 0.99\}$}	
\includegraphics[width=.3\linewidth]{Figures/gaus.png}
\includegraphics[width=.3\linewidth]{Figures/shapebeta.png}
\includegraphics[width=.3\linewidth]{Figures/mixednorm.png}
\end{figure}

@@

\includegraphics[width=\linewidth]{Figures/gaus_mse.png}

\includegraphics[width=\linewidth]{Figures/shapebeta_mse.png}

\includegraphics[width=\linewidth]{Figures/mixednorm_mse.png}

\section{Peaks Simulation}
We use another simulation design based on the applied problem we aim to solve. We assume that the measured data can be represented by 
\begin{equation}
Y(t) = s(t) + b(t) + \epsilon
\end{equation} 
where $s(t)$ is the true signal at time $t$, $b(t)$ is the drift component that varies smoothly over time and $\epsilon \sim N(0, \sigma^2)$ is an error component. We assume $t$ is a uniformly spaced sequence between 0 and 1. We generate $b(t)$ using a cubic natural spline basis function with degrees of freedom sampled from 2 to 10 with equal probability, and coefficients drawn from a normal distribution with mean and variance equal to 1. The true signal function is assumed to be zero with Gaussian peaks. The number of peaks is sampled from 5 to 15 with equal probability with centers uniformly distributed between 0.1 and 0.9 and bandwidths uniformly distributed between $2/n$ and $2/n + .01$. One hundred datasets were generated for $n=500$ and $n=1000$.

>>>>

\subsection*{Methods}
We compare the performance of our quantile trend filtering method with the three previously published methods using designs proposed by \cite{Racine2017}. The methods compared are

@@

	\item \texttt{detrendr\_SIC}: Our method where we minimize
	$\sum_i\rho_{\tau}(y_i - \theta_i) + \lambda||D\theta||_1$ and $\lambda$ is chosen using SIC from above. A single value of $\lambda$ was chosen by scaling and summing SIC values across all quantiles. 
	\item \texttt{detrendr\_valid}: Our method where lambda is chosen by leaving out every 5th observation as a validation data set and evaluating the check loss function on the validation data.
	\item \texttt{detrendr\_eBIC}:   The traditional BIC is given by 
	\begin{equation}
	\mbox{BIC}(s) = -2\log(L\{\hat{\theta}(s)\}) + \nu(s)\log n 
	\end{equation}	
	where $\theta(s)$ is the parameter $\theta$ with those components outside $s$ being set to 0, and $\nu(s)$ is the number of components in $s$. If we assume an asymmetric Laplace likelihood $L(y|\theta) = \left(\frac{\tau^n(1-\tau)}{\sigma}\right)^n\exp\left\{-\sum_i\rho_\tau(\frac{y_i - \theta_i}{\sigma})\right\}$ and the number of non-zero elements of $D\theta$ as $df$
	\begin{equation}
	\mbox{BIC}(df) = 2\sum_i\frac{1}{\sigma}\rho_{\tau}(y_i-\theta_i) + df\log n
	\end{equation} 
	We can choose and $\sigma>0$ and have found empirically that $\sigma =  \frac{1-|1-2\tau|}{2}$ produces stable estimates. \cite{chen2008} proposed the extended BIC for large parameter spaces 
	\begin{equation}
	BIC_{\gamma}(s) = -2\log(L\{\hat{\theta}(s)\}) + \nu(s)\log n  + 2\gamma\log{P\choose j,}~~\gamma \in [0,1]
	\end{equation}
	where $P$ is the total number of possible parameters and $j$ is the number of parameters included in given model. We used this criteria with $\gamma = 1$, $P=n-k$ where $k$ is the order of the differencing matrix and $j = \nu(s)$ is the number of non-zero entries in $D^{(k)}\theta$.

@@

\subsection*{BIC examples}

We used a single dataset to illustrate the difference between the scaled, unscaled and extended BIC criteria.

\begin{figure}[h!]
	\caption{Degrees of freedom (number of non-zero elements of $D\theta$) by $\log(\lambda)$.} 
	\includegraphics[width = 0.25\linewidth]{Figures/BIC_data.png}
	\includegraphics[width = 0.25\linewidth]{Figures/df_by_lambda.png}
	\includegraphics[width = 0.5\linewidth]{Figures/BIC_by_lambda.png}
\end{figure}

\subsection*{Design}
Three simulation designs from \cite{Racine2017} were considered. For all designs $X_i$ was generated as a uniformly spaced sequence in $[0,1]$ and the response $Y$ was generated as 
$$Y_i = sin(2\pi x_i) + \epsilon_i(x_i)$$
The three error distributions considered were 
\begin{itemize}
 \item Gaussian: $\epsilon_i(x_i) \sim N\left(0, \left(\frac{1+x_i^2}{4}\right)^2\right)$
 \item Beta: $\epsilon_i \sim Beta(1, 11-10x_i)$
\item Mixed normal: $\epsilon_i$ is simulated from a mixture of $N(-1,1)$ and  $N(1,1)$ with mixing probability $x_i$.
\end{itemize}
\begin{figure}
\caption{Simulated data with true quantiles $\tau \in \{0.01, 0.05, 0.25, 0.5, .75, 0.95, 0.99\}$}	
\includegraphics[width=.3\linewidth]{Figures/gaus.png}
\includegraphics[width=.3\linewidth]{Figures/shapebeta.png}
\includegraphics[width=.3\linewidth]{Figures/mixednorm.png}
\end{figure}

@@

\includegraphics[width=.6\linewidth]{Figures/gaus_mse.png}

\includegraphics[width=.6\linewidth]{Figures/shapebeta_mse.png}

\includegraphics[width=.6\linewidth]{Figures/mixednorm_mse.png}

\section{Peaks Simulation}
We use another simulation design based on the applied problem we aim to solve. We assume that the measured data can be represented by 
\begin{equation}
Y(t) = s(t) + b(t) + \epsilon
\end{equation} 
where $s(t)$ is the true signal at time $t$, $b(t)$ is the drift component that varies smoothly over time and $\epsilon \sim N(0, \sigma^2)$ is an error component. We assume $t$ is a uniformly spaced sequence between 0 and 1. We generate $b(t)$ using a cubic natural spline basis function with degrees of freedom sampled from 2 to 10 with equal probability, and coefficients drawn from a normal distribution with mean and variance equal to 1. The true signal function is assumed to be zero with Gaussian peaks. The number of peaks is sampled from 5 to 15 with equal probability with centers uniformly distributed between 0.1 and 0.9 and bandwidths uniformly distributed between $2/n$ and $2/n + .01$. One hundred datasets were generated for $n=\{300,500,1000\}$.
